Think deep for reef resilience

### 2025-07-28 paper/paper_draft.qmd
Updated paper abstract and conclusion paragraphs

---
title: Think deep for reef resilience
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
    attributes:
        corresponding: true
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier Reef (GBR) experiencing severe declines in coral cover from thermal stress events. Understanding which reefs exhibit resilience to climate impacts is critical for conservation planning and intervention strategies. A scenario-based population modelling approach was adopted with larval connectivity dynamics, thermal adaptation, and environmental factors to assess coral reef resilience across GBR locations under future climate projections. Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities, three distinct reef trajectories are identified: resilient reefs maintaining coral cover through 2050, intermediate reefs with moderate declines, and vulnerable reefs experiencing severe degradation. Contrary to expectations, larval connectivity patterns did not substantially differentiate resilient from vulnerable reefs. Instead, reef depth emerged as the primary predictor of resilience, with deeper reefs benefiting from thermal refugia that reduced bleaching impacts. Initial coral cover also influenced trajectories, with reefs having lower initial cover showing greater relative resilience. Our findings suggest that depth provides critical protection against thermal stress, contrasting current restoration practices that primarily target easily accessible shallow sites. These results have important implications for reef management, indicating that depth refugia should be prioritized for further consideration in conservation planning alongside connectivity considerations to maximize coral survival under continuing climate change.
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
editor:
  markdown:
    wrap: 92
---

@@

Statistically downscaled 1km resolution Sea Surface Temperature (SST) data for the GBR were
obtained from the HighResCoralStress database [@dixon2022; @dixon2023]. Maximum monthly mean
DHW levels were calculated from the SST data following NOAA methodology [@liu2017;
@skirving2020]. DHWs were represented by a selection of five Global Climate Models (GCMs)
from the Sixth Phase of the Coupled Model Intercomparison Project (CMIP6; **citation**).
These five GCMs were EC-Earth3-Veg, ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and GDFL-CM4, and
their inclusion were based on evaluation by Grose et al. [-@grose2023]. Grose et al.
[-@grose2023] found these GCMs to rank highly, among 50 GCMs, in recreation of historical
atmospheric temperature and precipitation conditions in the Australasian region. While this
assessment does not consider oceanic data, Grose et al. \[\@-grose2023\] provide a
region-specific selection mechanism which we use as an indicator of general performance of
GCMs for the region.

@@

# Conclusions

Overall, our results suggest that when modelled as part of a large and complex ecosystem,
reef depth is a key factor contributing to ongoing reef resilience. We provide evidence in
support of depth as a form of refugia for shallow reefs. While protecting ‘key node’ reefs
in larval connectivity networks is an important consideration for reef restoration efforts
[@anthony2020; @hock2017], our results suggest that current connectivity modelling methods
are not able to generate connectivity levels that contribute to reef resilience under
projected conditions. We suggest further analyses that investigate the benefits of
incorporating depth-based resilience into reef decision making, as well as further research
and development in connectivity modelling methods for whole-ecosystem scales.

>>>>

---
title: Think deep for reef resilience
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
    attributes:
        corresponding: true
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier 
    Reef (GBR) experiencing severe declines in coral cover from thermal stress events. 
    Understanding which reefs exhibit resilience to climate impacts is critical for conservation 
    planning and intervention strategies. A scenario-based population modelling approach was 
    adopted with larval connectivity dynamics, thermal adaptation, and environmental factors 
    to assess coral reef resilience across GBR locations under future climate projections. 
    Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities, 
    we analysed reef cover timeseries and the ability of reefs to maintain positive carbonate 
    production budgets under future conditions. Contrary to expectations, larval connectivity 
    patterns did not correlate with increases in maintainence of positive carbonate budgets. 
    Instead, reef depth emerged as the primary predictor of cover-timeseries resilience and 
    maintainence of positive carbonate budgets, with deeper reefs benefiting from thermal 
    refugia that reduce bleaching impacts. Our findings suggest that depth provides critical 
    protection against thermal stress, contrasting current restoration practices that primarily 
    target easily accessible shallow sites. These results have important implications for 
    reef management, indicating that depth refugia should be prioritized for further 
    consideration in conservation planning alongside connectivity considerations to maximize 
    coral survival under continuing climate change.
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
editor:
  markdown:
    wrap: 92
---

@@

Statistically downscaled 1km resolution Sea Surface Temperature (SST) data for the GBR were
obtained from the HighResCoralStress database [@dixon2022; @dixon2023]. Maximum monthly mean
DHW levels were calculated from the SST data following NOAA methodology [@liu2017;
@skirving2020]. DHWs were represented by a selection of five Global Climate Models (GCMs)
from the Sixth Phase of the Coupled Model Intercomparison Project (CMIP6). These five GCMs
were EC-Earth3-Veg, ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and GDFL-CM4, and their inclusion
were based on evaluation by Grose et al. [-@grose2023]. Grose et al. [-@grose2023] found
these GCMs to rank highly, among 50 GCMs, in recreation of historical atmospheric
temperature and precipitation conditions in the Australasian region. While this assessment
does not consider oceanic data, Grose et al. \[\@-grose2023\] provide a region-specific
selection mechanism which we use as an indicator of general performance of GCMs for the
region.

@@

In conclusion, our results suggest that when modelled as part of a large and complex
ecosystem, reef depth is a key factor contributing to ongoing reef resilience. We provide
evidence in support of depth as a form of thermal refugia for shallow-water reefs on the
GBR. While protecting ‘key node’ reefs in larval connectivity networks is an important
consideration for reef restoration efforts [@anthony2020; @hock2017], our results suggest
that current connectivity modelling methods are not able to generate connectivity that
contributes to reef resilience under projected conditions. We suggest further analyses that
investigate the benefits of incorporating depth-based resilience into reef decision making,
as well as further research and development in connectivity modelling methods for
whole-ecosystem scales.
